---
title: "Canadian Minimum Stipend Data Analysis"
output:
  html_document:
    df_print: paged
---
```{r data_import}
read.csv("SOS_ Minimum Stipend Data Collection - Finalized Information - Machine.csv")->sos

#us Data Source: https://www.nature.com/articles/d41586-022-01392-w
read.csv("Biology Benchmark Comparison - 07-23-2024.csv")->us.data
  #exchange rate
  us.to.cad<-1.38
  #convert from usd to cad internally below


#uk https://www.ukri.org/apply-for-funding/studentships-and-doctoral-training/london-weighting-enhanced-stipends-and-co-funding/
  uk.stipend<-19237
  uk.to.cad<-1.78
  #convert from gbp to cad
  uk.stipend*uk.to.cad->uk.stipend

#NSERC Scholarships
  NSERC.msc<-27000
  NSERC.phd<-40000
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lme4)
library(lmerTest)
library(ggplot2)
library(visreg)
```

#Functions
```{r}
select = dplyr::select #necessary as MASS also has a select function
filter = dplyr::filter

#Standard error of the mean
std <- function(x) sd(x)/sqrt(length(x))


scale_this <- function(x){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}

```

### NOTES from during converstion to a machine readable datafile ###
McMaster, both departments, provide "more to offset higher tuition" to the international MSc students, but the amount is unspecified.


# Summary information
There were `r length(sos$University)` programs within the final dataset, from `r length(unique(sos$University))` Universities. There are `r length(which(sos$Program == "MSc"))` MSc and `r length(which(sos$Program == "PhD"))` PhD programs. Of all of these different degree programs, `r length(which(sos$Example.Department == "Biological Sciences - Ecology"))/length(sos$University)*100`% were classified as closest to Ecology, while `r length(which(sos$Example.Department == "Physics"))/length(sos$University)*100`% are in Physics.

Mean domestic tuition was `r mean(sos$Domestic.Tuition)`CAD, median `r median(sos$Domestic.Tuition)`CAD. Mean international tuition was `r mean(sos$International.Tuition)`CAD, median `r median(sos$International.Tuition)`CAD. Mean domestic fees were `r mean(sos$Ancillary.Fees,na.rm=TRUE)`CAD, international mean `r mean(sos$Int.Ancillary.Fees,na.rm=TRUE)`CAD. Total tuition and fees domestic mean is `r mean(sos$Total.Domestic.Fees)`CAD, while international total was `r mean(sos$Total.Int.Fees,na.rm=TRUE)`CAD. 

Of the `r length(sos$University)` programs within the final dataset, `r length(which(sos$Guaranteed.Funding == "Yes"))` programs have a minimum, while `r length(which(sos$Guaranteed.Funding == ""))` had no information provided that we could access and did not respond to our requests. The average domestic minimum stipend provided was `r mean(sos$Gross.Stipend,na.rm=TRUE)`CAD, median `r median(sos$Gross.Stipend,na.rm=TRUE)`CAD. _When support is different than domestic support_, international minimum gross stipend mean is `r mean(sos$Gross.Stipend.International,na.rm=TRUE)`CAD. Net domestic mean is `r mean(sos$Net.Stipend,na.rm=TRUE)`CAD while median `r median(sos$Net.Stipend,na.rm=TRUE)`CAD. 

Mean tuition Get score: `r mean(sos$Tuition.Get...2,na.rm=T)`
Mean tuition Parse score: `r mean(sos$Tuition.Parse...2,na.rm=T)`
Mean Stipend Get score: `r mean(sos$Stipend.Get...3,na.rm=T)`
Mean Stipend Complete score: `r mean(sos$Stipend.Complete...3,na.rm=T)`

Domestic students repay `r mean(sos$Total.Domestic.Fees[which(sos$Guaranteed.Funding == "Yes")]/sos$Gross.Stipend[which(sos$Guaranteed.Funding == "Yes")]*100,na.rm=T)`% back to the university in the form of tuition and fees.
International students repay `r mean(sos$Total.Int.Fees[which(sos$Guaranteed.Funding == "Yes")]/sos$Gross.Stipend.International[which(sos$Guaranteed.Funding == "Yes")]*100,na.rm=T)`% back to the university in the form of tuition and fees.



Mean MBM Shortfall Domestic: `r mean(sos$Net.domestic...MBM.Threshold..2023.[which(sos$Guaranteed.Funding == "Yes")],na.rm=T)`
Mean MBM Shortfall International: `r mean(sos$Net.International...MBM.Threshold..2023.,na.rm=T)`

difference between US and Canada: `r mean(sos$Net.Stipend[which(sos$Guaranteed.Funding == "Yes")]) - mean(us.data$Annual.guaranteed.salary*us.to.cad)`
```{r visual_inspection}
sos.df <- as.data.frame(sos)
data.class(sos.df$Gross.Stipend)
data.class(sos.df$Net.Stipend)
sos.df$endowment <- sos.df$University.Endowment..Millions.of.Dollars.
data.class(sos.df$endowment)

#Histogram of values - visually inspect to see if skewed
histGross <- hist(sos.df$Gross.Stipend,main = "Gross Stipend",xlab = "Value ($CAD)")
histNet <- hist(sos.df$Net.Stipend,main = "Net Stipend",xlab = "Value ($CAD)")

summary(sos.df$Gross.Stipend)
summary(sos.df$Net.Stipend)

#Remove NAs
sos.df <- sos.df %>%
  dplyr::filter(!is.na(Gross.Stipend))%>%
  dplyr::filter(!is.na(Net.Stipend))

summary(sos.df$Gross.Stipend)
summary(sos.df$Net.Stipend)

#Histogram of values
histGross <- hist(sos.df$Gross.Stipend,main = "Gross Stipend",xlab = "Value ($CAD)")
histNet <- hist(sos.df$Net.Stipend,main = "Net Stipend",xlab = "Value ($CAD)")

#Scale stipend 
sos.df<- sos.df %>%
  mutate(scaledGrossStipend = scale(Gross.Stipend, center = TRUE, scale = TRUE), 
         scaledNetStipend = scale(Net.Stipend, center = TRUE, scale = TRUE))

summary(sos.df$scaledGrossStipend)
sd(sos.df$scaledGrossStipend)

summary(sos.df$scaledNetStipend)
sd(sos.df$scaledNetStipend)


#Histogram of values - visually inspect to see if skewed
histGrossscaled <- hist(sos.df$scaledGrossStipend,main = "Gross Stipend",xlab = "Value ($CAD)")
histNetscaled <- hist(sos.df$scaledNetStipend,main = "Net Stipend",xlab = "Value ($CAD)")



#Endowment
histEndow <- hist(sos.df$endowment,main = "Endowment",xlab = "Value ($CAD)") #highly skewed

#Log-transform 
sos.df<- sos.df %>%
  mutate(logEndowment = log10(endowment)) 

histlogEndow <- hist(sos.df$logEndowment,main = "Endowment (log10 transformed)",xlab = "Value ($CAD)") 
```

#Plots for the paper.
```{r paper_plots}
# PhD vs. MSc
  par(mfcol=c(1,2))
      #setting up a stable y to maintain through various plots
        ylimits<-c(min(c(sos$Net.domestic...MBM.Threshold..2024.,
                         {us.data$Annual.guaranteed.salary-us.data$MIT.Living.Wage..1.adult..0.children.}*us.to.cad,
                         uk.stipend-uk.cost.of.living.min,
                         uk.stipend-uk.cost.of.living.max), 
              na.rm=T),
                   max(c(sos$Gross.Stipend,
                         {us.data$Annual.guaranteed.salary}*us.to.cad,
                         uk.stipend,
                         uk.stipend), 
              na.rm=T)
                   )
        
  ## Gross Stipend (Domestic) Boxplots        
    boxplot(sos$Gross.Stipend[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Program[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Gross Domestic")
    abline(h=0,col="black")
    abline(h=c(NSERC.msc,NSERC.phd),col="red")
    
    boxplot(us.data$Annual.guaranteed.salary*us.to.cad,
            ylim=ylimits,
            main="US [box],UK [line]")
    abline(h=uk.stipend,col="blue")
    abline(h=0,col="black")
```
Left panel is Canadian domestic students gross minimum stipends, two red are for federally supported students, higher is PhD, lower MSc. Right is for international comparisons. Blue is UK, which provides a single stipend amount (exception: London) for the entire funding system.
    
```{r Net_domestic}      
  ## Net Stipend (Domestic) Boxplots
    boxplot(sos$Net.Stipend[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Program[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Net Domestic")
    abline(h=0,col="black")
    boxplot(us.data$Annual.guaranteed.salary*us.to.cad,
            ylim=ylimits,
            main="US,UK")
    abline(h=0,col="black")
    abline(h=uk.stipend,col="blue")
```
Same labeling conventions as above.
```{r domestic_poverty}
  ## Vs. poverty threshold (Domestic)
  par(mfcol=c(1,1))

    boxplot(sos$Net.domestic...MBM.Threshold..2023.[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Program[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Supported Domestic students vs. poverty line")
        abline(h=0)
```
Same labeling conventions as above. This has had the Market Basket Metric removed, as a measure of poverty provided by the government. Again, this is only for students with support. All students in departments without a minimum graduate stipend have been removed.
## Supplemental figures
```{r domestic_poverty_Rentals}
  ## Vs. poverty threshold (Domestic)
  par(mfcol=c(1,1))

    boxplot(sos$Net.domestic...Rent.year[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Program[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Supported Domestic students after paying rent")
        abline(h=0)
```
This is just after paying an average rent for their area and tuition and fees. No food, for example, is included.
```{r domestic_poverty_Rentals_food}
  ## Vs. poverty threshold (Domestic)
  par(mfcol=c(1,1))

    boxplot(sos$Net.domestic...Rent....12K[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Program[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Supported Domestic students after paying rent and assuming ~$12k for food & incidentals")
        abline(h=0)
```





```{r University funding }

plot(sos$University.Endowment..Millions.of.Dollars.[which(sos$Example.Department == "Physics")],
     sos$Net.Stipend[which(sos$Example.Department == "Physics")],
     pch=16,
     col="purple",
     ylim=c(0,max(sos$Net.Stipend,na.rm=T)),
     main="Univ. Endowment vs. Net Stipends")
points(sos$University.Endowment..Millions.of.Dollars.[which(sos$Example.Department == "Biological Sciences - Ecology")],
     sos$Net.Stipend[which(sos$Example.Department == "Biological Sciences - Ecology")],
     pch=15,
     col="forestgreen")
cor.test(sos$University.Endowment..Millions.of.Dollars.,
     sos$Net.Stipend,use="complete.obs")

# Line of correlation
  lm(sos$Net.Stipend[which(sos$Example.Department == "Biological Sciences - Ecology")] ~ sos$University.Endowment..Millions.of.Dollars.[which(sos$Example.Department == "Biological Sciences - Ecology")] )->lm.end
  abline(lm.end)
  
#chi sq tests
    #median(sos$University.Endowment..Millions.of.Dollars.,na.rm=T)->ue
    #median(sos$Net.Stipend,na.rm=T)->ns
  # or 
    1500->ue
    15000->ns
abline(v=ue)
abline(h=ns)

length(which(sos$University.Endowment..Millions.of.Dollars. >= ue & 
               sos$Net.Stipend >= ns))->up.right
length(which(sos$University.Endowment..Millions.of.Dollars. < ue & 
               sos$Net.Stipend >= ns))->up.left
length(which(sos$University.Endowment..Millions.of.Dollars. < ue & 
               sos$Net.Stipend < ns))->down.left
length(which(sos$University.Endowment..Millions.of.Dollars. >= ue & 
               sos$Net.Stipend < ns))->down.right
# table
as.table(rbind(c(up.left,up.right),c(down.left,down.right)))->table
chisq.test(table)




#Q-Q Plots for scaled Gross Stipend

  data.class(sos.df$Gross.Stipend)
  sos.df$Gross.Stipend.SQ <- sqrt(sos.df$Gross.Stipend)
  sos.df$Gross.Stipend.SQ4 <- sos.df$Gross.Stipend^(0.25)
  sos.df$Gross.Stipend.Log <- log(sos.df$Gross.Stipend)+1
  
  
  op<-par(mfrow=c(2,2), mar=c(4,3,3,2))
  qqnorm(sos.df$Gross.Stipend, main="None (p=1)")
  qqline(sos.df$Gross.Stipend)
  qqnorm(sos.df$Gross.Stipend.SQ, main="Square root (p=0.5)")
  qqline(sos.df$Gross.Stipend.SQ)
  qqnorm(sos.df$Gross.Stipend.SQ4,main="Fourth root (p=0.25)")
  qqline(sos.df$Gross.Stipend.SQ4)
  qqnorm(sos.df$Gross.Stipend.Log,main="Logarirthmic (p=0)")
  qqline(sos.df$Gross.Stipend.Log)
  par(op)


#Q-Q Plots for scaled Net Stipend
  
  data.class(sos.df$Net.Stipend)
  sos.df$Net.Stipend.SQ <- sqrt(sos.df$Net.Stipend)
  sos.df$Net.Stipend.SQ4 <- sos.df$Net.Stipend^(0.25)
  sos.df$Net.Stipend.Log <- log(sos.df$Net.Stipend)+1
  
  
  op<-par(mfrow=c(2,2), mar=c(4,3,3,2))
  qqnorm(sos.df$Net.Stipend, main="None (p=1)")
  qqline(sos.df$Net.Stipend)
  qqnorm(sos.df$Net.Stipend.SQ, main="Square root (p=0.5)")
  qqline(sos.df$Net.Stipend.SQ)
  qqnorm(sos.df$Net.Stipend.SQ4,main="Fourth root (p=0.25)")
  qqline(sos.df$Net.Stipend.SQ4)
  qqnorm(sos.df$Net.Stipend.Log,main="Logarirthmic (p=0)")
  qqline(sos.df$Net.Stipend.Log)
  par(op)



#Filter NA endowment data 
sos.df <- sos.df %>%
    filter(!is.na(endowment)) %>%
    collect()%>%
    filter(Guaranteed.Funding == "Yes")

#Number of data rows remaining
  nrow(sos.df) 
  n_distinct(sos.df$University)
  
  sos.datasummary <- sos.df %>%
  group_by(Province) %>% 
    summarise(nrecords = length (University)) 
  sos.datasummary


#GLMM Net
  NMSm1 <- lmer(Net.Stipend ~ 
                    logEndowment
                    + Program
                    + Example.Department
                    + (1|Province/University),   data = sos.df)
  NMSm1
  
  summary(NMSm1)
  visreg(NMSm1, "logEndowment", by = "Program")
  visreg(NMSm1, "logEndowment", by = "Example.Department")
  visreg(NMSm1, "logEndowment", by = "Province")
  resid(NMSm1)
  plot(density(resid(NMSm1)))
  qqnorm(resid(NMSm1))
  plot(NMSm1)
  AIC(NMSm1)
  plot(sos.df$endowment,resid(NMSm1))


#Fitted vs. Residual w Smooth line
  plot(NMSm1, type=c("p","smooth"), col.line=1)
#Scale-Location
  plot(NMSm1,
       sqrt(abs(resid(.)))~fitted(.),
       type=c("p","smooth"), col.line=1)
#Residuals v Leveragae
  plot(NMSm1, rstudent(.) ~ hatvalues(.))

```
# OLS Fit for endowment
```{r OLS Fit}
# Perform OLS regression
model <- lm(Net.Stipend ~ Program + logEndowment, data = sos.df)

# View the summary of the regression
summary(model)

# Plot 1: Residuals vs logEndowment
plot(sos.df$logEndowment, residuals(model),
     xlab = "Log Endowment", ylab = "Residuals",
     main = "Residuals vs Log Endowment")
abline(h = 0, col = "red", lty = 2)

# Create a scatter plot with the regression line
ggplot(sos.df, aes(x = logEndowment, y = Net.Stipend)) +
  geom_point(color = "blue") +  # Plot the data points
  geom_smooth(method = "lm", formula = y ~ x, color = "red") +  # Add the regression line
  labs(title = "Regression of Net.Stipend on logEndowment",
       x = "log(Endowment)",
       y = "Net Stipend") 

plot(model)



```

```{r Annual_expenses}
plot(sos$University.Total.Annual.Expenses.2023..Millions.of.Dollars.[which(sos$Example.Department == "Physics")],
     sos$Net.Stipend[which(sos$Example.Department == "Physics")],
     pch=16,
     col="purple",
     ylim=c(0,max(sos$Net.Stipend,na.rm=T)),
     main="Annual Expenses vs. Net stipends")
points(sos$University.Total.Annual.Expenses.2023..Millions.of.Dollars.[which(sos$Example.Department == "Biological Sciences - Ecology")],
     sos$Net.Stipend[which(sos$Example.Department == "Biological Sciences - Ecology")],
     pch=15,
     col="forestgreen")
cor.test(sos$University.Total.Annual.Expenses.2023..Millions.of.Dollars.,
     sos$Net.Stipend,use="complete.obs")


```
Biplots comparing the Universities endowment size vs. the funding received by the student in the first plot, while the second plot compares the annual expenses of the university against the net stipend. We are using the annual expenses as a rough proxy for the size of the university.

# Field comparisons
```{r field_comparisons}

## Vs. field (Domestic)
  par(mfcol=c(1,2))

    boxplot(sos$Net.Stipend[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Example.Department[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="net domestic stipend, department")
        abline(h=0)
        
    boxplot(sos$Net.domestic...MBM.Threshold..2023.[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Example.Department[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="MBM Shortfall, department")
        abline(h=0)
```
Comparisons between Biology and Physics, in the same style as the above. Here just net domestic stipend and the MBM Shortfall. 

# Provincial Comparisons
```{r provincial}
#Province
  ## Gross Stipend (Domestic) Boxplots
    boxplot(sos$Gross.Stipend[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Province[which(sos$Guaranteed.Funding == "Yes")],
        ylim=ylimits,
        main="Gross provincial")
    abline(h=0)
  ## Net Stipend (Domestic) Boxplots  
    boxplot(sos$Net.Stipend[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Province[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Net Provincial")
    abline(h=0)
  ## Vs. poverty threshold (Domestic)
    boxplot(sos$Net.domestic...MBM.Threshold..2023.[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Province[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="MBM Shortfall, provincial")
    abline(h=0)
```

# International Plots
```{r International}
mean(sos$Total.Int.Fees[which(sos$Guaranteed.Funding == "Yes")]/sos$Gross.Stipend.International[which(sos$Guaranteed.Funding == "Yes")]*100,
       na.rm=T)
median(sos$Total.Int.Fees[which(sos$Guaranteed.Funding == "Yes")]/sos$Gross.Stipend.International[which(sos$Guaranteed.Funding == "Yes")]*100,
       na.rm=T)

  par(mfcol=c(1,1))

    boxplot(sos$Net.International...MBM.Threshold..2023.[which(sos$Guaranteed.Funding == "Yes")] ~ sos$Program[which(sos$Guaranteed.Funding == "Yes")],
            ylim=ylimits,
            main="Supported International students vs. poverty line")
        abline(h=0)

```
